---
title: "Exoplanets project - new "
output: html_document
date: "2025-03-17"
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Categories

Classification Criteria:

Sub-Earth:
  Radius: < 0.5 Earth radii
  Mass: < 0.5 Earth masses
  Temperature: 150 - 400K
  Small, likely rocky planets.

Earth-like:
  Radius: 0.5 - 1.5 Earth radii
  Mass: 0.5 - 5 Earth masses
  Temperature: 180 - 370K
  Likely to be rocky, similar to Earth.

Cold Super-Earth:
  Radius: 1.5 - 2.5 Earth radii
  Mass: 5 - 10 Earth masses
  Temperature: 100 - 180K
  Larger than Earth but still possibly rocky.

Hot Super-Earth:
  Radius: 1.5 - 2.5 Earth radii
  Mass: 5 - 10 Earth masses
  Temperature: 370 - 800K
  Larger than Earth but still possibly rocky.

Mini-Neptune:
  Radius: 2.5 - 4 Earth radii
  Mass: 10 - 50 Earth masses
  Temperature: 300 - 1000K
  Likely to have thick atmospheres.

Neptune-like:
  Radius: 4 - 6 Earth radii
  Mass: 50 - 100 Earth masses
  Temperature: 500 - 1500K
  Gas-dominated, similar to Neptune.

Gas Giant:
  Radius: > 6 Earth radii
  Mass: > 100 Earth masses
  Temperature: > 800K
  Similar to Jupiter & Saturn.

Then we added the comparison with their Equlibrium Temperatures and Host Star Temperatures, to see whether they would be viable parameters.

# Basic understanding the data 

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)

# Load dataset - need to change the path
data <- read.csv("C:/Users/krali/Downloads/nasa_classified.csv")

# Define planet types based on radius, mass, and temperature

classify_planet <- function(radius, mass, temp) {
  if (is.na(radius) | is.na(mass) | is.na(temp)) {
    return("Unknown")  # Assign "Unknown" if any value is NA
  } else if (radius < 0.5 & mass < 0.5 & temp >= 150 & temp <= 400) {
    return("Sub-Earth")
  } else if (radius >= 0.5 & radius <= 1.5 & mass >= 0.5 & mass <= 5 & temp >= 180 & temp <= 370) {
    return("Earth-like")
  } else if (radius > 1.5 & radius <= 2.5 & mass > 5 & mass <= 10 & temp >= 100 & temp < 180) {
    return("Cold Super-Earth")
  } else if (radius > 1.5 & radius <= 2.5 & mass > 5 & mass <= 10 & temp > 370 & temp <= 800) {
    return("Hot Super-Earth")
  } else if (radius > 2.5 & radius <= 4 & mass > 10 & mass <= 50 & temp > 300 & temp <= 1000) {
    return("Mini-Neptune")
  } else if (radius > 4 & radius <= 6 & mass > 50 & mass <= 100 & temp > 500 & temp <= 1500) {
    return("Neptune-like")
  } else if (radius > 6 & mass > 100 & temp > 800) {
    return("Gas Giant")
  } else {
    return("Unknown")
  }
}
# Apply classification function
data$Planet_Type <- mapply(classify_planet, data$pl_rade, data$pl_bmasse, data$pl_eqt)

# 1. Bar Chart: Number of Exoplanets by Type
planet_counts <- data %>% group_by(Planet_Type) %>% summarise(count = n())

ggplot(planet_counts, aes(x = Planet_Type, y = count, fill = Planet_Type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Sub-Earth"= "darkgreen", "Earth-like" = "green","Hot Super-Earth"="pink", "Cold Super-Earth" = "blue", "Mini-Neptune" = "purple", "Neptune-like" = "orange", "Gas Giant" = "red", "Unknown" = "gray")) +
  labs(title = "Number of Exoplanets by Type", x = "Planet Type", y = "Count") +
  theme_minimal()

planet_counts <- data %>%
  group_by(Planet_Type) %>%
  summarise(count = n()) %>%
  filter(Planet_Type != "Unknown")  # Exclude "Unknown" category

# the same but without unknown
ggplot(planet_counts, aes(x = Planet_Type, y = count, fill = Planet_Type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Sub-Earth"= "darkgreen", "Earth-like" = "green","Hot Super-Earth"="pink", "Cold Super-Earth" = "blue", "Mini-Neptune" = "purple", "Neptune-like" = "orange", "Gas Giant" = "red")) +
  labs(title = "Number of Exoplanets by Type", x = "Planet Type", y = "Count") +
  theme_minimal()

# 2. Boxplot: Temperature Distributions by Planet Type
ggplot(data, aes(x = Planet_Type, y = pl_eqt, fill = Planet_Type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Sub-Earth"= "darkgreen", "Earth-like" = "green","Hot Super-Earth"="pink", "Cold Super-Earth" = "blue", "Mini-Neptune" = "purple", "Neptune-like" = "orange", "Gas Giant" = "red", "Unknown" = "gray")) +
  labs(title = "Equilibrium Temperature by Planet Type", x = "Planet Type", y = "Equilibrium Temperature (K)") +
  theme_minimal()

# 3. Scatter Plot: Mass vs. Radius with filtred data so it is visible
# for planets with radius < 10
filtered_data <- data %>% filter(pl_rade < 10)

ggplot(filtered_data, aes(x = pl_rade, y = pl_bmasse, color = Planet_Type)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("Sub-Earth"= "darkgreen", "Earth-like" = "green","Hot Super-Earth"="pink", "Cold Super-Earth" = "blue", "Mini-Neptune" = "purple", "Neptune-like" = "orange", "Gas Giant" = "red", "Unknown" = "gray")) +
  labs(title = "Exoplanet Mass vs. Radius (Radius < 10 Earth radii)", 
       x = "Radius (Earth radii)", 
       y = "Mass (Earth masses)") +
  theme_minimal()

# the same but without the uknown planets
filtered_data <- data %>% filter(pl_rade < 10 & pl_bmasse < 150 & Planet_Type != "Unknown")

ggplot(filtered_data, aes(x = pl_rade, y = pl_bmasse, color = Planet_Type)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("Sub-Earth"= "darkgreen", "Earth-like" = "green","Hot Super-Earth"="pink", "Cold Super-Earth" = "blue", "Mini-Neptune" = "purple", "Neptune-like" = "orange", "Gas Giant" = "red", "Unknown" = "gray")) +
  labs(title = "Exoplanet Mass vs. Radius (Radius < 10 Earth radii)", 
       x = "Radius (Earth radii)", 
       y = "Mass (Earth masses)") +
  theme_minimal()

```

### Focus on the orbital periods 

```{r}
#Code with orbital periods

# Filter data with orbital period < 600 days and radius < 10 Earth radii otherwise the plot is not good to read
filtered_data <- data %>% filter(pl_orbper < 600, pl_rade < 10, Planet_Type != "Unknown")

# 1. Scatter Plot: Orbital Period vs. Planet Radius
ggplot(filtered_data, aes(x = pl_rade, y = pl_orbper, color = Planet_Type)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("Sub-Earth"= "darkgreen", "Earth-like" = "green","Hot Super-Earth"="pink", "Cold Super-Earth" = "blue", "Mini-Neptune" = "purple", "Neptune-like" = "orange", "Gas Giant" = "red", "Unknown" = "gray")) +
  labs(title = "Orbital Period vs. Planet Radius", 
       x = "Planet Radius (Earth radii)", 
       y = "Orbital Period (days)") +
  theme_minimal()

# 2. Boxplot: Orbital Period by Planet Type
ggplot(filtered_data, aes(x = Planet_Type, y = pl_orbper, fill = Planet_Type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Sub-Earth"= "darkgreen", "Earth-like" = "green","Hot Super-Earth"="pink", "Cold Super-Earth" = "blue", "Mini-Neptune" = "purple", "Neptune-like" = "orange", "Gas Giant" = "red", "Unknown" = "gray")) +
  labs(title = "Orbital Period by Planet Type", 
       x = "Planet Type", 
       y = "Orbital Period (days)") +
  theme_minimal()

# 3. Bubble Chart: Orbital Period vs. Planet Mass
ggplot(filtered_data, aes(x = pl_orbper, y = pl_bmasse, size = pl_rade, color = Planet_Type)) +
  geom_point(alpha = 0.6) +
  scale_size_continuous(range = c(2, 8)) +
  scale_color_manual(values = c("Sub-Earth"= "darkgreen", "Earth-like" = "green","Hot Super-Earth"="pink", "Cold Super-Earth" = "blue", "Mini-Neptune" = "purple", "Neptune-like" = "orange", "Gas Giant" = "red", "Unknown" = "gray")) +
  labs(title = "Bubble Chart: Orbital Period vs. Planet Mass (filtred)", 
       x = "Orbital Period (days)", 
       y = "Planet Mass (Earth masses)", 
       size = "Planet Radius (Earth radii)") +
  theme_minimal()

# 4. Bar Chart: Number of Planets by Orbital Period Range
data$orbital_period_range <- cut(data$pl_orbper, 
                                 breaks = c(0, 10, 50, 100, 200, 500, 1000, Inf), 
                                 labels = c("< 10 days", "10-50 days", "50-100 days", "100-200 days", "200-500 days", "500-1000 days", "> 1000 days"), 
                                 right = FALSE)

planet_counts_orbit <- data %>% group_by(orbital_period_range) %>% summarise(count = n())

# 5. Plot the bar chart without x-axis labels
ggplot(planet_counts_orbit, aes(x = orbital_period_range, y = count, fill = orbital_period_range)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Number of Planets by Orbital Period Range (Unknown Excluded)", 
       x = "Orbital Period Range", 
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_blank())  # Remove x-axis labels


```

### Distance to the star graphs 
We dont have the information about the earth-like planets so this can be used for the other categories.

```{r}
# Boxplot: Distance to the Closest Star by Planet Type
ggplot(filtered_data, aes(x = Planet_Type, y = sy_dist, fill = Planet_Type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Sub-Earth"= "darkgreen", "Earth-like" = "green","Hot Super-Earth"="pink", "Cold Super-Earth" = "blue", "Mini-Neptune" = "purple", "Neptune-like" = "orange", "Gas Giant" = "red")) +
  labs(title = "Distance to Closest Star by Planet Type", 
       x = "Planet Type", 
       y = "Distance to Closest Star (pc)") +
  theme_minimal()

# Bubble Chart: Distance to Closest Star vs. Planet Mass
ggplot(filtered_data, aes(x = sy_dist, y = pl_bmasse, size = pl_rade, color = Planet_Type)) +
  geom_point(alpha = 0.6) +
  scale_size_continuous(range = c(2, 8)) +
  scale_color_manual(values = c("Sub-Earth"= "darkgreen", "Earth-like" = "green","Hot Super-Earth"="pink", "Cold Super-Earth" = "blue", "Mini-Neptune" = "purple", "Neptune-like" = "orange", "Gas Giant" = "red")) +
  labs(title = "Bubble Chart: Distance to Closest Star vs. Planet Mass", 
       x = "Distance to Closest Star (pc)", 
       y = "Planet Mass (Earth masses)", 
       size = "Planet Radius (Earth radii)") +
  theme_minimal()



```

## Understanding the planets that are earth-like


```{r}

# Print the number of Earth-like exoplanets
earth_like_count <- planet_counts$count[planet_counts$Planet_Type == "Earth-like"]
print(paste("Number of Earth-like exoplanets:", earth_like_count))

# Filter the dataset to get Earth-like planets
earth_like_planets <- filtered_data %>% filter(Planet_Type == "Earth-like")

ggplot(earth_like_planets, aes(x = pl_eqt, y = pl_bmasse, size = pl_rade)) +
  geom_point(alpha = 0.7, color = "darkblue") +
  labs(title = "Bubble Chart: Radius vs. Mass (250K ≤ Temperature ≤ 350K)",
       x = "Equilibrium Temperature (K)",
       y = "Planet Mass (Me)",
       size = "  Planet Radius (Re)") +
  scale_size_continuous(range = c(2, 8)) +
  theme_minimal()


# View the table of Earth-like planets
print(earth_like_planets)

```


Because all 4 of our earth like planets are in 1 solar system with 1 star so we can compare that one to the star

```{r}
library(dplyr)

# Create a subset of data for Sun and TRAPPIST-1
sun_data <- data.frame(
  star = "Sun",
  st_teff = 5778,        # Sun's effective temperature in Kelvin
  st_rad = 1,            # Sun's radius in solar radii
  st_mass = 1,           # Sun's mass in solar masses
  st_lum = 1,            # Sun's luminosity in solar units
  st_met = 0.014,        # Sun's metallicity
  st_logg = 4.44         # Sun's surface gravity in log units
)

# TRAPPIST-1 data (You may need to pull the actual data from your dataset if available)
trappist_data <- data %>%
  filter(hostname == "TRAPPIST-1") %>%
  select(st_teff, st_rad, st_mass, st_lum, st_met, st_logg) %>%
  distinct() %>%
  mutate(star = "TRAPPIST-1")

# Combine Sun and TRAPPIST-1 data
comparison_data <- bind_rows(sun_data, trappist_data)

# Display the comparison data
comparison_data

```


# Interactive window for information about planets in same categories 

```{r}
library(shiny)
library(plotly)
library(tidyr)

ui <- fluidPage(
  titlePanel("🚀 Exoplanet Comparison Dashboard"),
  sidebarLayout(
    sidebarPanel(
      selectInput(
        "selected_category", 
        "🌍 Select Planet Category:", 
        choices = c("All", "Earth-like", "Super-Earth", "Mini-Neptune", "Neptune-like", "Gas Giant", "Unknown"),
        selected = "All"
      ),
      selectInput(
        "selected_planets", 
        "🪐 Select Exoplanets:", 
        choices = NULL,
        multiple = TRUE
      ),
      actionButton("add_earth", "🌎 Add Earth for Comparison")  
    ),
    mainPanel(
      plotlyOutput("radar_chart"),
      plotlyOutput("temperature_chart"),
      plotlyOutput("orbital_period_chart"),  
      plotlyOutput("mass_temperature_chart"),
      uiOutput("planet_info")  
    )
  )
)

server <- function(input, output, session) {
  
  earth_added <- reactiveVal(FALSE)  
  
  observeEvent(input$selected_category, {
    df <- data %>%
      mutate(Planet_Type = mapply(classify_planet, pl_rade, pl_bmasse, pl_eqt))
    
    if (input$selected_category != "All") {
      df <- df %>% filter(Planet_Type == input$selected_category)
    }
    
    updateSelectInput(session, "selected_planets", choices = unique(df$pl_name))
  })
  
  filtered_data <- reactive({
    req(input$selected_planets)
    df <- data %>%
      mutate(Planet_Type = mapply(classify_planet, pl_rade, pl_bmasse, pl_eqt))
    
    if (input$selected_category != "All") {
      df <- df %>% filter(Planet_Type == input$selected_category)
    }
    
    df %>% filter(pl_name %in% input$selected_planets)
  })
  
  output$radar_chart <- renderPlotly({
    df <- filtered_data() %>% select(pl_name, st_rad, pl_orbsmax, pl_rade, pl_bmasse)
    df_long <- tidyr::pivot_longer(df, cols = c(st_rad, pl_orbsmax, pl_rade, pl_bmasse), names_to = "Variable", values_to = "Value")
    
    if (earth_added()) {
      earth_long <- data.frame(
        pl_name = "Earth", 
        Variable = c("st_rad", "pl_orbsmax", "pl_rade", "pl_bmasse"),
        Value = c(1, 1, 1, 1)
      )
      df_long <- rbind(df_long, earth_long)
    }

    plot_ly(df_long, r = ~Value, theta = ~Variable, color = ~pl_name, type = "scatterpolar", mode = "lines+markers", fill = "toself") %>%
      layout(
        title = "📊 Radar Chart: Exoplanet and Stellar Features",
        polar = list(radialaxis = list(visible = TRUE, range = c(0, max(df_long$Value))))
      )
  })
  
  output$temperature_chart <- renderPlotly({
    df <- filtered_data()
    
    if (earth_added()) {
      earth_data <- df[1, ]
      earth_data$pl_name <- "Earth"
      earth_data$pl_eqt <- 288
      earth_data[, !names(earth_data) %in% c("pl_name", "pl_eqt")] <- NA
      df <- rbind(df, earth_data)
    }

    plot_ly(df, x = ~pl_name, y = ~pl_eqt, type = "bar", color = ~pl_name) %>%
      layout(title = "🌡️ Exoplanet Temperature Comparison", yaxis = list(title = "Temperature (K)"))
  })
  
  output$orbital_period_chart <- renderPlotly({
    df <- filtered_data()
    
    # if (earth_added()) {
    #  earth_data <- df[1, ]
    #   earth_data$pl_name <- "Earth"
    #   earth_data$pl_orbper <- 365
    #   earth_data[, !names(earth_data) %in% c("pl_name", "pl_orbper")] <- NA
    #   df <- rbind(df, earth_data)
    # }

    plot_ly(df, x = ~pl_name, y = ~pl_orbper, type = "bar", color = ~pl_name) %>%
      layout(title = "🌀 Orbital Period Comparison", yaxis = list(title = "Orbital Period (Days)"))
  })
  
  output$planet_info <- renderUI({
    df <- filtered_data()
    planet_info <- lapply(1:nrow(df), function(i) {
      planet <- df[i, ]
      tagList(
       h4(paste0("🪐 ", planet$pl_name)),
      p(strong("🌟 Star System:"), planet$hostname),
      p(strong("🛸 Mass:"), planet$pl_masse, " Earth masses"),
      p(strong("🔵 Radius:"), planet$pl_rade, " Earth radii"),
      p(strong("🔥 Equilibrium Temperature:"), planet$pl_eqt, "K"),
      p(strong("☀️ Insolation:"), planet$pl_insol),
      p(strong("⏳ Orbital Period:"), planet$pl_orbper, " days"),
      p(strong("⚖️ Density:"), planet$pl_dens, " g/cm³"), 
      p(strong("💫 Orbital Eccentricity:"), planet$pl_orbeccen),  
      hr()
      )
    })
    do.call(tagList, planet_info)
  })
  
  observeEvent(input$add_earth, {
    earth_added(!earth_added())  
  })
}

shinyApp(ui, server)

```

## Interactive window for Metallicity Visualization

```{r}
# UI
ui <- fluidPage(
  titlePanel("Interactive Metallicity Visualization"),
  
  sidebarLayout(
    sidebarPanel(
      selectizeInput("planet_type", "Select Planet Type:", 
               choices = unique(data$Planet_Type), 
               multiple = TRUE,
               options = list(placeholder = 'Select a planet type')),
      radioButtons("x_axis", "X-axis:", choices = c("Radius (pl_rade)" = "pl_rade", 
                                                     "Mass (pl_bmasse)" = "pl_bmasse"), selected = "pl_rade")
    ),
    
    mainPanel(
      tabsetPanel(
        tabPanel("Scatter Plot", plotlyOutput("scatterPlot")),
        tabPanel("Metallicity Histogram", plotlyOutput("histogram"))
      )
    )
  )
)

# Server
server <- function(input, output) {
  
  filtered_data <- reactive({
    data %>% filter(Planet_Type %in% input$planet_type)
  })
  
  output$scatterPlot <- renderPlotly({
    p <- ggplot(filtered_data(), aes_string(x = input$x_axis, y = "st_met", color = "Planet_Type")) +
      geom_point(alpha = 0.7, size = 3) +
      labs(x = input$x_axis, y = "Metallicity (st_met)", title = "Metallicity vs. Radius/Mass") +
      theme_minimal()
    
    ggplotly(p)
  })
  
  output$histogram <- renderPlotly({
    p <- ggplot(filtered_data(), aes(x = st_met, fill = Planet_Type)) +
      geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
      labs(x = "Metallicity (st_met)", y = "Count", title = "Distribution of Metallicity") +
      theme_minimal()
    
    ggplotly(p)
  })
}

# Run the application
shinyApp(ui = ui, server = server)
```
